---
title: "Incident Trends on NYC Subway Lines During and Post-COVID"
editor: visual
---
The COVID-19 pandemic profoundly impacted urban public transportation systems worldwide, and New York City’s subway network was no exception. The Metropolitan Transportation Authority (MTA) faced unprecedented challenges as ridership plummeted, operational demands shifted, and new patterns of incidents emerged during and after the pandemic. Understanding these changes is critical for enhancing the resilience and safety of this vital transit system.
This report explores the most common categories of incidents reported on New York City subway lines during and after the COVID-19 pandemic, spanning from 2020 to 2024. Leveraging a comprehensive dataset from the NYC Open Data platform, the analysis examines incidents categorized by line, division, month, day type, and delay category. Through this granular examination, the report identifies key trends, highlights significant operational challenges, and provides insights into shifting patterns of subway incidents over time.

## Overview of Incident Trends
```{r echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true
#| code-summary: "Show the code"

# Install required packages if not already installed
if (!require("tidytext")) install.packages("tidytext")
library(tidytext)

if (!require("RColorBrewer")) install.packages("RColorBrewer")
library(RColorBrewer)

if (!require("lubridate")) install.packages("lubridate")
library(lubridate)

if (!require("stringr")) install.packages("stringr")
library(stringr)

if (!require("gt")) install.packages("gt")
library(gt)

if (!require("dplyr")) install.packages("dplyr")
library(dplyr)

if (!require("ggplot2")) install.packages("ggplot2")
library(ggplot2)

if (!require("tidyverse")) install.packages("tidyverse")
library(tidyverse)

if (!require("DT")) install.packages("DT")
library(DT)

if (!require("plotly")) install.packages("plotly")
library(plotly)

# Function to load data
mtaincident_file <- function(filepath) {
  as.data.frame(readr::read_csv(filepath, lazy = FALSE, show_col_types = FALSE))
}

# File path
path <- "C:\\Users\\w3038\\Downloads\\STA 9750\\Final Project\\MTA_Subway_Major_Incidents__Beginning_2020_20241026.csv"

# Load data
mtaincident <- mtaincident_file(path)

# Clean the data by removing empty rows in 'line' column
mtaincident <- mtaincident %>%
  filter(line != "", line != "Systemwide") %>%
  mutate(division = case_when(
    division == "A DIVISION" ~ "Number Line",
    division == "B DIVISION" ~ "Letter Line",
    TRUE ~ division  # Retain other values
  ))

# Convert the 'month' column to Date format
mtaincident$month <- as.Date(mtaincident$month, format = "%m/%d/%Y")

# Aggregate the data to get total complaints by month and category
monthly_category_Incidents <- mtaincident %>%
  group_by(month, category) %>%
  summarise(total_Incidents = sum(count)) %>%
  ungroup()

# Create a static stacked area chart using ggplot
static_area_chart <- ggplot(monthly_category_Incidents, aes(x = month, y = total_Incidents, fill = category)) +
  geom_area(alpha = 0.8, color = "black", size = 0.2) +
  labs(title = "Monthly Incidents by Category Over Time",
       x = "Month",
       y = "Total Incidents",
       fill = "Incidents Category") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set3") +
  theme(plot.title = element_text(hjust = 0.5))

# Convert the static chart to an interactive chart using plotly
interactive_area_chart <- ggplotly(static_area_chart)

# Display the interactive chart
interactive_area_chart

```
From 2020 to 2024, NYC subway incidents exhibit a noticeable upward trend across all categories. An interactive stacked area chart will visualize the distribution of incidents by category over time, illustrating the dynamics of subway operations. Among the categories, incidents related to "persons on trackbed/police/medical" consistently represent the highest counts, posing a significant challenge for the MTA. Other high-frequency categories include signals, track, and subway car-related incidents. In contrast, incidents categorized under stations and structure show relatively lower counts, indicating fewer disruptions in these areas.

The upward trend in incidents, particularly during the post-COVID period, underscores the critical need for targeted interventions and resource allocation. The interactive chart will allow users to isolate specific trends, hover over data points to view incident counts, and explore seasonal or monthly patterns. This functionality enhances the depth of analysis and supports data-driven decision-making for improving subway performance and safety.


## Comparative Analysis: COVID vs. Post-COVID Incident Distribution
```{r echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true
#| code-summary: "Show the code"
library(scales)

# Load and clean MTA incident data
mtaincident$month <- as.Date(mtaincident$month)  # Ensure month column is in Date format


# Calculate total incidents for each category during the COVID period (2020-2021)
covid_incident_reasons <- mtaincident %>%
  filter(month >= as.Date("2020-01-01") & month < as.Date("2022-01-01")) %>%
  group_by(category) %>%
  summarise(total_incidents = sum(count, na.rm = TRUE)) %>%
  mutate(percentage = total_incidents / sum(total_incidents) * 100)

# Plot pie chart for COVID period
covid_pie <- ggplot(covid_incident_reasons, aes(x = "", y = total_incidents, fill = category)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  geom_text(aes(label = paste0(round(percentage, 1), "%")), 
            position = position_stack(vjust = 0.5), color = "white") +
  labs(title = "Incident Reasons During COVID Period (2020-2021)", fill = "Incident Category") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5))

# Display COVID pie chart
print(covid_pie)

# Calculate total incidents for each category during the post-COVID period (2022-2023)
post_covid_incident_reasons <- mtaincident %>%
  filter(month >= as.Date("2022-01-01") & month < as.Date("2024-01-01")) %>%
  group_by(category) %>%
  summarise(total_incidents = sum(count, na.rm = TRUE)) %>%
  mutate(percentage = total_incidents / sum(total_incidents) * 100)

# Plot pie chart for post-COVID period
post_covid_pie <- ggplot(post_covid_incident_reasons, aes(x = "", y = total_incidents, fill = category)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  geom_text(aes(label = paste0(round(percentage, 1), "%")), 
            position = position_stack(vjust = 0.5), color = "white") +
  labs(title = "Incident Reasons Post-COVID Period (2022-2023)", fill = "Incident Category") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5))

# Display post-COVID pie chart
print(post_covid_pie)

```

To ensure a balanced analysis, both the COVID (2020-2021) and post-COVID (2022-2023) periods are scaled to two years each. Pie charts will illustrate the distribution of incident categories for each period. During the COVID period, signal-related incidents were most prevalent, comprising 33.2% of all incidents, followed by persons on trackbed/police/medical incidents at 30.2%. Other notable categories included track-related incidents (10.8%), subway car incidents (8.9%), and a smaller share for stations and structure (5.7%). These distributions highlight the operational challenges of maintaining reliability and addressing immediate safety concerns during the pandemic.

Post-COVID, the distribution shifts significantly. Persons on trackbed/police/medical incidents rose to 33.6%, becoming the largest category, while signal-related incidents decreased to 24.6%. Track-related incidents surged to 18.2%, reflecting emerging infrastructure maintenance challenges. Categories such as subway car (8.0%), stations and structure (8.8%), and other (6.8%) incidents showed relatively smaller changes. These shifts indicate evolving priorities for MTA as the system transitions to a recovery phase, emphasizing the need for focused interventions to address rising safety and infrastructure issues.

## Incident Trends by Subway Line
```{r echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true
#| code-summary: "Show the code"

# Load and clean MTA incident data
mtaincident$month <- as.Date(mtaincident$month)  # Ensure month column is in Date format

# Filter data for COVID (2020-2021) and post-COVID (2022-2023) periods
covid_data <- mtaincident %>%
  filter(month >= as.Date("2020-01-01") & month < as.Date("2022-01-01"))

post_covid_data <- mtaincident %>%
  filter(month >= as.Date("2022-01-01") & month < as.Date("2024-01-01"))

# Calculate total incidents by line and division for COVID period
line_incident_covid <- covid_data %>%
  group_by(line, division) %>%
  summarise(total_incident = sum(count, na.rm = TRUE)) %>%
  arrange(desc(total_incident))

# Calculate total incidents by line and division for post-COVID period
line_incident_post <- post_covid_data %>%
  group_by(line, division) %>%
  summarise(total_incident = sum(count, na.rm = TRUE)) %>%
  arrange(desc(total_incident))

# Combine COVID and post-COVID datasets for mirrored plotting
line_incidents_combined <- bind_rows(
  line_incident_covid %>% mutate(period = "COVID (2020-2021)"),
  line_incident_post %>% mutate(period = "Post-COVID (2022-2023)")
)

# Separate data by division
number_line_data <- line_incidents_combined %>%
  filter(division == "Number Line")

letter_line_data <- line_incidents_combined %>%
  filter(division == "Letter Line")

# Create mirrored chart for Number Line
number_line_graph <- ggplot(number_line_data, aes(x = reorder(line, total_incident), y = ifelse(period == "COVID (2020-2021)", -total_incident, total_incident), fill = period)) +
  geom_bar(stat = "identity", position = "identity") +
  geom_text(aes(label = abs(total_incident)), 
            position = position_stack(vjust = 0.5), 
            color = "black", 
            size = 3.5) +  # Display total incidents as text
  scale_y_continuous(labels = abs) +  # Show absolute values on y-axis
  labs(title = "Total Incidents by Number Line (COVID vs. Post-COVID)",
       x = "Subway Line",
       y = "Total Incidents") +
  theme_minimal() +
  coord_flip() +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_fill_manual(values = c("lightblue", "lightgreen"))

# Create mirrored chart for Letter Line
letter_line_graph <- ggplot(letter_line_data, aes(x = reorder(line, total_incident), y = ifelse(period == "COVID (2020-2021)", -total_incident, total_incident), fill = period)) +
  geom_bar(stat = "identity", position = "identity") +
  geom_text(aes(label = abs(total_incident)), 
            position = position_stack(vjust = 0.5), 
            color = "black", 
            size = 3.5) +  # Display total incidents as text
  scale_y_continuous(labels = abs) +  # Show absolute values on y-axis
  labs(title = "Total Incidents by Letter Line (COVID vs. Post-COVID)",
       x = "Subway Line",
       y = "Total Incidents") +
  theme_minimal() +
  coord_flip() +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_fill_manual(values = c("lightblue", "lightgreen"))

# Display the plots
print(number_line_graph)
print(letter_line_graph)

```
A detailed line graph analysis compares incident trends on NYC subway lines during the COVID and post-COVID periods. Separate graphs for lettered and numbered lines reveal significant increases in total incidents post-COVID, suggesting a correlation with ridership recovery and operational adjustments.

### **Letter Subway Lines**:
- **Notable Increases**: Lines like N, E, and F show sharp rises in incidents post-COVID. For example, N line incidents increased from 69 during COVID to 85 post-COVID, while E and F lines recorded jumps of +35 and +20 incidents, respectively.
- **Minimal Changes**: S Franklin and S Rockaway lines experienced consistently low incidents, highlighting their relatively stable performance.

### **Number Subway Lines**:
- **Significant Rises**: The 2 line saw the largest increase among all lines, jumping from 48 incidents during COVID to 92 post-COVID (+44 incidents). The 6 line also recorded a substantial rise (+19 incidents), while the 1 line’s incidents nearly doubled from 30 to 58.
- **Low-Incident Outliers**: Lines like S 42nd St showed minimal change, maintaining very low incidents across both periods.
These findings suggest that the increase in incidents is linked to higher service activity and ridership levels post-pandemic. Lines with the highest increases, such as the 2 and N lines, may require additional attention to address operational and safety challenges. Conversely, consistently low-incident lines could offer insights into effective practices that can be replicated across the network.

## Key Insights and Implications
- **Dominance of “Persons on Trackbed/Police/Medical” Incidents**:
During the COVID-19 pandemic, signal-related incidents were the most prevalent (33.2%). However, post-COVID, the share of "persons on trackbed/police/medical" incidents increased significantly, becoming the largest category at 33.6%.
This shift underscores the growing safety and security concerns in the subway system, requiring targeted interventions.
- **Infrastructure and Maintenance Challenges Post-COVID**: Track-related incidents experienced a sharp rise, increasing from 10.8% during COVID to 18.2% post-COVID. This highlights emerging challenges in maintaining subway infrastructure as ridership recovered.
- **Consistent Trends Across Specific Categories**: Categories like subway car incidents remained relatively stable, with minor changes (8.9% during COVID vs. 8.0% post-COVID). Stations and structure-related incidents also stayed on the lower end of the spectrum, showing the least disruption compared to other categories.
- **Subway Line Variability**: Certain subway lines exhibited sharp increases in incidents post-COVID, particularly the N, E, and F letter lines and the 2 and 6 numbered lines.
Low-incident lines, such as S Franklin and S Rockaway, consistently maintained minimal disruptions, pointing to potential best practices for safety and operations.


## Conclusion

The analysis of NYC subway incidents during and post-COVID highlights significant shifts in incident types and their distribution. The rise in “persons on trackbed/police/medical” incidents post-pandemic underscores the need for enhanced safety protocols and preventive measures to address behavioral and security-related issues. Additionally, the surge in track-related incidents emphasizes the urgency of prioritizing infrastructure maintenance as ridership rebounds.

Subway lines with notable incident increases, such as the N, E, and 2, require focused attention to address operational challenges and improve reliability. Meanwhile, low-incident lines, like S Franklin and S 42nd St, could serve as benchmarks for replicating effective practices system-wide.

Understanding these trends enables the MTA to identify areas of concern, allocate resources strategically, and implement data-driven policies to improve safety and operational resilience. Addressing these evolving challenges will be critical in ensuring the reliability and safety of NYC’s subway system in the post-pandemic era.




